A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-14 text/fig_legend.Rmd
Update figures with 3d umap, fig ref in text, fig legend text

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
Classification of tumors by pathway disruptions.** a) Each of 377 selected
Reactome pathways (columns) is classified as disrupted if one or more genes is
mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
standardized abbreviations from the TCGA project. b) MCA analysis was performed to 
determine the eigenvalue with the most explanatory power from the categorical data.
Average of 100 permutations of the data was used as a baseline. c) The p-value of
each eigenvalue was plotted and the max eigenvalue with a pvalue < 0.05 was selected. 
d) The resulting MCA was projected down to 3 dimensions for the purpose of density 
based clustering (only the first 2 dimensions are shown). Each dot represents a tumor 
e) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize
the fraction of unassigned tumors. f) HDBSCAN clustered tumors were visualized on the UMAP
projection; tumors are colored by cluster membership (see legend). This results
in some tumors remaining unclassified (grey points). g) kNN was used to assign
membership to the remaining tumors. h) Mutational burden plot where each dot corresponds
to the sample mutation counts for tumors in each class. i) Projection of cancer 
type identity onto UMAP. Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma,
HNSC = Head and neck squamous cell, PAAD = Pancreatic adenocarcinoma. j) Projection of breast
cancer subtypes onto the UMAP. k) Projection of BRCA1/2 somatic mutation onto the
UMAP. l) Projection of tumor stage onto the UMAP, regardless of cancer type. m) 
Projection of metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, 
M1 = metastatic tumors.</p>

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 2: Pan-cancer 
enrichment of pathway disruptions.** a) Waterfall plot; columns represent tumors. 
Each row represents a pathway. The pathways are ranked such that the pathway with 
the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are
ordered to prioritize the top pathway disruptions successively. The top 100 most 
mutated pathways are shown. Red signifies pathway disrupted pathways, grey signifies
not disrupted. b) Heatmap shows the relative enrichment of each pathway (columns) 
within each numbered cluster (rows). Effect size is displayed as colors (orange, 
enrichment; purple, depletion). c) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size >= 0.30 in each cluster)
within each numbered cluster (rows).</p>

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3: Gene level 
analysis reveals tissue-specific class signatures.** Odds ratio plot; Column uses 
a logarithmic axis to represent odds ratio with a 95% confidence interval. Row 
represents the significant genes from each cluster. Each cluster was compared 
against the background (all other clusters) to find deferentially mutated genes. 
Significant genes (pvalue < 0.01) were selected and limited to the top ten results 
for each clusters. Heatmap; Column represents the cancer type. Row represents the 
significant genes as described above. The heatmap shows the proportion of samples 
per cluster, per cancer type that is mutated for each of the significant genes. 
Depleted significant genes (left). Enriched significant genes (right).</p>

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1: Cluster-specific 
enriched pathways (effect size >= 0.30) in metastatsis.**

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary figure S1. 
Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
in **Figure 1D of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S2. 
Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
cancers projected onto the tumors of the UMAP from **Figure 1D of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S3.
Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15  
genes involved in genome stability projected on the UMAP from **Figure 1D of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S4. 
Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
pathways projected on the UMAP from **Figure 1D of the main text** ordered by the 
hierarchical pathway order (column) from **Figure 2B of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S5.
Distribution of oncogenes on UMAP clusters.** The left panel shows the identity of 
clusters 1, 7 and 8. The multipanel on the right shows one of the three oncogenes 
projected onto the tumors of the UMAP from **Figure 1D of the main text**.

>>>>

**Figure 1: Molecular pathway profiles of tumor samples with one or more
mutations.** Each of 377 selected Reactome pathways (columns) is classified
as disrupted if one or more genes is mutated in the tumor sample (rows) where
red and grey represents the pathway is mutated and not, respectively. Tumors
are grouped by tissue of origin using standardized abbreviations 
from the TCGA project.

**Figure 2: Method for tumor classification using pathway aberration data** 
a) MCA analysis was performed to determine the eigenvalue with the most 
explanatory power from the categorical data. Average of 100 permutations of 
the data was used as a baseline. b) The p-value of each eigenvalue was plotted
and the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA 
was projected down to 3 dimensions for the purpose of density based clustering 
(only the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN
was used to perform hierarchical density based clustering tuned to minimize
the fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized 
on the UMAP projection; tumors are colored by cluster membership (see legend). 
This results in some tumors remaining unclassified (grey points). f) kNN was 
used to assign membership to the remaining tumors.

**Figure 3: Pathway-based classification is independent of tissue-of-origin features.**
a) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer, 
COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD = 
Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the UMAP.
c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of tumor 
stage onto the UMAP, regardless of cancer type. e) Projection of metastatic status
onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.

**Figure 4: Somatic mutation frequencies for tumors in each class** Column uses 
a logarithmic axis to represent mutation count. Row represents the tumor samples. 
Each dot corresponds to the sample mutation count for tumors in each class, 
ordered by mutation counts. The median mutation count in each cluster is indicated 
by the horizontal line.

**Figure 5: Pan-cancer enrichment of pathway disruptions.** a) Waterfall plot; columns
represent tumors. Each row represents a pathway. The pathways are ranked such that the 
pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and 
tumors are ordered to prioritize the top pathway disruptions successively. The top 100 
most mutated pathways are shown. Red signifies pathway disrupted pathways, grey signifies
not disrupted. b) Heatmap shows the relative enrichment of each pathway (columns) 
within each numbered cluster (rows). Effect size is displayed as colors (orange, 
enrichment; purple, depletion). c) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size >= 0.30 in each cluster)
within each numbered cluster (rows).

**Figure 6: Gene level analysis reveals tissue-specific class signatures.** Odds ratio 
plot; Column uses a logarithmic axis to represent odds ratio with a 95% confidence 
interval. Row represents the significant genes from each cluster. Each cluster was 
compared against the background (all other clusters) to find deferentially mutated 
genes. Significant genes (pvalue < 0.01) were selected and limited to the top ten 
results for each clusters. Heatmap; Column represents the cancer type. Row represents the 
significant genes as described above. The heatmap shows the proportion of samples 
per cluster, per cancer type that is mutated for each of the significant genes. 
Depleted significant genes (left). Enriched significant genes (right).

**Figure 7: **

**Table 1: Cluster-specific enriched pathways (effect size >= 0.30) in metastatsis.**
